Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-05-20 draft.tex

Given the training set of synthetic light curves, SNN returns a classification (SN Ia, CCSN, or peculiar) for observed light curves. Before classifying, we remove transients with variability in multiple seasons (likely active galactic nuclei or superluminous SNe). SNN is then run on every transient with a host galaxy redshift (see Section \ref{subsubsec:sn_hosts}). We define SNe with a threshold probability of $P(\mathrm{Ia})\geq0.5$ as photometrically classified SNe Ia, although our analysis is not sensitive to this choice, since the vast majority of SNe receive classifications close to 1 or 0. The photometrically classified SNe Ia are then passed through a \texttt{SALT2} light curve fitting code and are subject to quality cuts on $x_1$ and $c$ in an identical manner to \citet{Vincenzi2020}: $-3 \leq x_1 \leq 3$, $-0.3 \leq c \leq 0.3$. This selection helps to reduce the potential contamination from core-collapse SNe (CCSNe) and outlying thermonuclear SNe, as well as removing those in extremely dusty environments. We impose further cuts on the quality of those measured parameters, which are: the uncertainty on $x_1$: $\sigma_{x_1} <1$; the uncertainty on the date at which the light curve has peak brightness: $\sigma_{t_{\mathrm{peak}}} <2$; the SALT2 fit probability\footnote{based on light curve fit $\chi^2$ and number of degrees of freedom.} $>0.01$. The number of SNe passing each stage of these cuts is displayed in Table 2 of \citet{Vincenzi2020} and reduces the $\sim30,000$ objects to 1604 SNe. With the addition of the photometric classifier we are left with 1441 SNe with which we proceed to inspect their host galaxies. While the inclusion of stringent cuts significantly reduces the size of the sample, it greatly enhances the purity and is in line with previous rates analyses of this nature \citep{Sullivan2006} as well as being consistent with other DES-SN analyses \citep[e.g.][]{Kelsey2021}. The contamination of this sample from CCSNe according to the simulations of \citet{Vincenzi2020} is expected to be below 3\%, which is close to that of spectroscopic samples \citep{Rubin2015}.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/SFHs_colour.png}
    \caption{The stellar mass assembly of the Universe as prescribed by the toy model of \citetalias{Childress2014}. Galaxies with high stellar mass at $z=0$ formed the majority of their stars in the distant past, while lower-mass galaxies are still actively star forming in the present day.}
    \label{fig:SFHs}
\end{figure*}

@@

\begin{equation}
    R_{\mathrm{G}} = \int_{t_0}^{\infty} \psi(t_0-\tau)\Phi(\tau)\,d\tau \,,
    \label{eq:galaxy_rate}
\end{equation}
where $t_0$ is the epoch at which the galaxy is observed, and $\psi$ is the SFH.

@@

\renewcommand{\arraystretch}{1.2}
\begin{table}
	\centering
	\caption{Results of the Bayesian parameter estimation for the SN Ia DTD.}
	\label{tab:dtd_results}
	\begin{tabular}{lccr} % four columns, alignment for each
		\hline
		 &$\beta$ & $A$ & $t_{\mathrm{p}}$\\
		 &-       & $10^{-13}$ M$_{\odot}^{-1}$yr$^{-1}$ & Gyr \\
		\hline
		Fixed $t_{\mathrm{p}}$ & $-1.14\pm0.05$ &  $2.08 \pm 0.08$ & -\\
		Free $t_{\mathrm{p}}$ & $-1.03 _{-0.07} ^{+0.11}$ & $1.87 _{-0.17} ^{+0.13}$ & $0.07 _{-0.04} ^{+0.03}$\\
		Fixed $A$, $\beta$ & - & - & $0.047_{-0.007}^{+0.008}$\\
		\hline
	\end{tabular}
\end{table}

@@

\subsubsection{The DTD power law index \label{subsubsec:compare_beta}}
We measure a DTD power law index of $-1.14\pm0.05$. This value is consistent with the majority of previous analyses using various methods. Our measurement provides an independent support to the values close to $-1$ found using volumetric rates \citep[e.g.][]{Graur2011,Graur2014,Frohmaier2019}, individual galaxy SFHs \citep[e.g.][]{Maoz2012,Graur2013}, and galaxy clusters \citep[e.g.][]{Maoz2010} although our slope is marginally inconsistent with that of \citet{Heringer2019}. The result is also qualitatively consistent with that of \citet{Strolger2020a} who fit an exponential function DTD rather than a power law. However, their best fit model indicated a higher SN rate at long delay times than a $\tau^{-1}$ DTD, which is opposite to the mild preference of our fit which is steeper and thus produces a smaller fraction of SNe at long delay times.

\subsubsection{The DTD normalisation \label{subsubsec:compare_A}}
We measure a DTD normalisation (i.e. the rate of SNe 1 Gyr after star-formation) of $1.73 \pm 0.05 \times 10^{-13}$~M$_{\odot}^{-1}$~yr$^{-1}$. \citet{Graur2013} found a value of $0.7 \times 10^{-13}$~M$_{\odot}^{-1}$~yr$^{-1}$ using the SFH technique, whereas \citet{Heringer2019} report $7 \times 10^{-13}$~M$_{\odot}^{-1}$~yr$^{-1}$. These results lie either side of our measured value. \citet{Heringer2019} note that the normalisation recovered from the SFHR method is sensitive to the assumed prompt time $t_{\mathrm{p}}$, a degeneracy that is also apparent our method. The normalisation is also sensitive to the assumed IMF which can explain this variation \citep{Maoz2017}.

@@

\renewcommand{\arraystretch}{1.2}
\begin{table}
	\centering
	\caption{Results of the Bayesian parameter estimation for the SN Ia DTD.}
	\label{tab:dtd_results_split_lc}
	\begin{tabular}{crr} % four columns, alignment for each
		\hline
		 Sample &$\beta$ & $A$ \\
		 &-       & $10^{-13}$ M$_{\odot}^{-1}$yr$^{-1}$ \\
		\hline
		Fiducial & $-1.14\pm0.05$ &  $2.08\pm0.08$ \\
	    $x_1 < 0$ & $-0.79\pm0.08$ &  $1.19 \pm0.05$ \\
		$x_1 > 0$ & $-1.68\pm0.15$ & $0.51\pm0.12$ \\
		$c < 0$ & $-1.09\pm0.08$ & $0.89\pm0.04$ \\
		$c > 0$ & $-1.19\pm0.09$ & $1.16\pm0.08$ \\
		\hline
	\end{tabular}
\end{table}
\subsection{Second order processes affecting the supernova rate \label{subsec:discussion}}

@@

One possible cause of the discrepant SN Ia rate at high stellar mass is the effect of stellar metallicity. Metallicity has been previously invoked to explain irregularities or divergences from the fiducial DTD \citep[e.g.][]{Strolger2010,Meng2011,Kistler2013}. Metallicity may affect the observed rate of SNe in two ways: firstly, it can affect the time taken for a star to evolve along the main sequence ($t_{\mathrm{MS}}$) -- Stellar metallicity has varying effects on the main-sequence lifetime of stars \citep[e.g.][]{Georgy2013,Amard2020}, which is also dependent upon initial rotation and degree of mixing; secondly it can affect the time taken from WD formation to SN Ia explosion -- low metallicity stars should produce higher-mass WDs \citep[e.g.][]{Umeda1999,Marigo2007}, resulting in a higher SN Ia rate \citep{Kistler2013}, although note that \citet{Kistler2013} did not find any evidence for this in the data. \citet{Graur2017a} also found that evolution of SN Ia rates with metallicity is consistent with a DTD of the form  $t^{-1}$.

@@

Alternatively, low- and high-stretch SNe could form via identical evolutionary channels but with a different onset time since the episode of star-formation. In this scenario, he predominantly high-stretch "prompt" SNe begin exploding as soon as the WDs are formed (nominally 40 Myr) whereas the mainly low-stretch "tardy" SNe only begin exploding after an extended period of time of the order 1 Gyr. The time dependence of both populations would follow the fiducial $t^{-1}$ distribution.
Such a scenario may be explained by the sub-$M_{\mathrm{Ch}}$ double-detonation paradigm \citep[e.g.][]{Sim2010,Blondin2017,Shen2017}. Sub-$M_{\mathrm{Ch}}$ SNe Ia typically involve the merger of two WDs and the SN luminosity can be correlated with the mass of the primary WD; a correlation between primary WD mass and age leads to different DTDs of low and high-stretch SNe and hence the different rates observed in this work. Such an evolution of light curve properties with age has recently been seen across the full range of "normal" SNe Ia in 1-dimensional simulations by \citet{Shen2021} and is a promising avenue for further investigation although we note that they struggle to reproduce observed light curves at later times ($t\gtrsim30$~d; \citealt{Shen2021,Gronow2021}).

@@

The SN Ia rates generated by the posterior predictive checks are shown in Fig. \ref{fig:rate_fitted_split_x1_2c_DTD}. The best fitting values are somewhat consistent with one another, with $t_1 = 0.65 \pm 0.1$~Gyr and $t_2 = 0.85 \pm 0.15$~Gyr, corresponding to a relatively sharp transition between the two populations. Our $t_1$ and $t_2$ are reminiscent of the transition found by \citet{Brandt2010} who found high-stretch SNe to be confined to delay times $\lesssim 0.4$~Gyr while low-stretch SNe occurred with delay times $\gtrsim 2.4$~Gyr. On the other hand, the transition found here appears to occur at much shorter delay times than the several Gyrs found by \citet{Rose2019}. These results are somewhat surprising, as we might expect that the transition is not so sharp but a gradual correlation of delay time with the average stretch of SNe caused by the age vs WD-mass distribution, as predicted by \citet{Shen2021}. However we note that the intrinsic stretch distribution may not be a simple Gaussian, but could instead be bimodal \citep{Scolnic2016,Popovic2021}. The correlation of progenitor age and WD mass does not easily reproduce such a bimodal distribution, which may suggest that the sub-$M_{\mathrm{Ch}}$ channel cannot account for all SNe Ia. We also note that our method models the average SFH of galaxies and that the DTD is a statistical distribution: our results do not rule out high-stretch SNe arising from old progenitors, but suggest that such a scenario is unlikely.

>>>>

Given the training set of synthetic light curves, SNN returns a classification (SN Ia, CCSN, or peculiar) for observed light curves. Before classifying, we remove transients with variability in multiple seasons (likely active galactic nuclei or superluminous SNe). SNN is then run on every transient with a host galaxy redshift (see Section \ref{subsubsec:sn_hosts}). We define SNe with a threshold probability of $P(\mathrm{Ia})\geq0.5$ as photometrically classified SNe Ia, although our analysis is not sensitive to this choice, since the vast majority of SNe receive classifications close to 1 or 0. The photometrically classified SNe Ia are then passed through a \texttt{SALT2} light curve fitting code and are subject to quality cuts on $x_1$ and $c$ in an identical manner to \citet{Vincenzi2020}: $-3 \leq x_1 \leq 3$, $-0.3 \leq c \leq 0.3$. This selection helps to reduce the potential contamination from core-collapse SNe (CCSNe) and outlying thermonuclear SNe, as well as removing those in extremely dusty environments. We impose further cuts on the quality of those measured parameters, which are: the uncertainty on $x_1$: $\sigma_{x_1} <1$; the uncertainty on the date at which the light curve has peak brightness: $\sigma_{t_{\mathrm{peak}}} <2$~d; the SALT2 fit probability\footnote{based on light curve fit $\chi^2$ and number of degrees of freedom.} $>0.01$. The number of SNe passing each stage of these cuts is displayed in Table 2 of \citet{Vincenzi2020} and reduces the $\sim30,000$ objects to 1604 SNe. With the addition of the photometric classifier we are left with 1441 SNe with which we proceed to inspect their host galaxies. While the inclusion of stringent cuts significantly reduces the size of the sample, it greatly enhances the purity and is in line with previous rates analyses of this nature \citep{Sullivan2006} as well as being consistent with other DES-SN analyses \citep[e.g.][]{Kelsey2021}. The contamination of this sample from CCSNe according to the simulations of \citet{Vincenzi2020} is expected to be below 3\%, which is close to that of spectroscopic samples \citep{Rubin2015}.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/SFHs_colour.png}
    \caption{The stellar mass assembly of the Universe as prescribed by the toy model of \citetalias{Childress2014}. The $y$-axis is the fraction of a galaxy's total mass of formed stars that are formed in each time bin. Galaxies with high stellar mass at $z=0$ formed the majority of their stars in the distant past, while lower-mass galaxies are still actively star forming in the present day.}
    \label{fig:SFHs}
\end{figure*}

@@

\begin{equation}
    R_{\mathrm{G}} = \int_{t_0}^{t_f} \psi(t_0-\tau)\Phi(\tau)\,d\tau \,,
    \label{eq:galaxy_rate}
\end{equation}
where $t_0$ is the epoch at which the galaxy is observed, $t_f$ is the time at which the first stars in the galaxy formed, and $\psi$ is the SFH.

@@

\renewcommand{\arraystretch}{1.2}
\begin{table}
	\centering
	\caption{Results of the Bayesian parameter estimation for the SN Ia DTD.}
	\label{tab:dtd_results}
	\begin{tabular}{lccr} % four columns, alignment for each
		\hline
		 &$\beta$ & $A$ & $t_{\mathrm{p}}$\\
		 &-       & $10^{-13}$ M$_{\odot}^{-1}$yr$^{-1}$ & Gyr \\
		\hline
		Fixed $t_{\mathrm{p}}$ & $-1.13^{+0.04}_{-0.06}$ &  $2.11^{+0.05}_{-0.12}$ & -\\
		%Free $t_{\mathrm{p}}$ & $-1.03 _{-0.07} ^{+0.11}$ & $1.87 _{-0.17} ^{+0.13}$ & $0.07 _{-0.04} ^{+0.03}$\\
		Fixed $A$, $\beta$ & - & - & $0.047_{-0.007}^{+0.008}$\\
		\hline
	\end{tabular}
\end{table}

@@

\subsubsection{The DTD power law index \label{subsubsec:compare_beta}}
We measure a DTD power law index of $-1.13^{+0.04}_{-0.06}$. This value is consistent with the majority of previous analyses using various methods. Our measurement provides an independent support to the values close to $-1$ found using volumetric rates \citep[e.g.][]{Graur2011,Graur2014,Frohmaier2019}, individual galaxy SFHs \citep[e.g.][]{Maoz2012,Graur2013}, and galaxy clusters \citep[e.g.][]{Maoz2010} although our slope is marginally inconsistent with that of \citet{Heringer2019}. The result is also qualitatively consistent with that of \citet{Strolger2020a} who fit an exponential function DTD rather than a power law. However, their best fit model indicated a higher SN rate at long delay times than a $\tau^{-1}$ DTD, which is opposite to the mild preference of our fit which is steeper and thus produces a smaller fraction of SNe at long delay times.

\subsubsection{The DTD normalisation \label{subsubsec:compare_A}}
We measure a DTD normalisation (i.e. the rate of SNe 1 Gyr after star-formation) of $2.11^{+0.05}_{-0.12} \times 10^{-13}$~M$_{\odot}^{-1}$~yr$^{-1}$. \citet{Graur2013} found a value of $0.7 \times 10^{-13}$~M$_{\odot}^{-1}$~yr$^{-1}$ using the SFH technique, whereas \citet{Heringer2019} report $7 \times 10^{-13}$~M$_{\odot}^{-1}$~yr$^{-1}$. These results lie either side of our measured value. \citet{Heringer2019} note that the normalisation recovered from the SFHR method is degenerate with the assumed prompt time $t_{\mathrm{p}}$, a degeneracy that is also reflected in the non-convergence of our fit that included $t_{\mathrm{p}}$ as a free parameter. The normalisation is also sensitive to the assumed IMF which can explain this variation \citep{Maoz2017}.

@@

\renewcommand{\arraystretch}{1.2}
\begin{table}
	\centering
	\caption{Results of the Bayesian parameter estimation for the SN Ia DTD.}
	\label{tab:dtd_results_split_lc}
	\begin{tabular}{crr} % four columns, alignment for each
		\hline
		 Sample &$\beta$ & $A$ \\
		 &-       & $10^{-13}$ M$_{\odot}^{-1}$yr$^{-1}$ \\
		\hline
		Fiducial & $-1.13\pm0.05$ &  $2.11\pm0.08$ \\
	    $x_1 < 0$ & $-0.79\pm0.08$ &  $1.19 \pm0.05$ \\
		$x_1 > 0$ & $-1.70^{+0.19}_{-0.10}$ & $0.50^{+0.26}_{-0.09}$ \\
		$c < 0$ & $-1.11^{+0.05}_{-0.11}$ & $0.90\pm0.04$ \\
		$c > 0$ & $-1.18\pm0.09$ & $1.14^{+0.11}_{-0.05}$ \\
		\hline
	\end{tabular}
\end{table}
\subsection{Second order processes affecting the supernova rate \label{subsec:discussion}}

@@

One possible cause of the discrepant SN Ia rate at high stellar mass is the effect of stellar metallicity. Metallicity has been previously invoked to explain irregularities or divergences from the fiducial DTD \citep[e.g.][]{Strolger2010,Meng2011,Kistler2013}. Metallicity may affect the observed rate of SNe in two ways: firstly, it can affect the time taken for a star to evolve along the main sequence ($t_{\mathrm{MS}}$) -- Stellar metallicity has varying effects on the main-sequence lifetime of stars \citep[e.g.][]{Georgy2013,Amard2020}, which is also dependent upon initial rotation and degree of mixing; secondly it can affect the time taken from WD formation to SN Ia explosion -- low metallicity stars should produce higher-mass WDs \citep[e.g.][]{Umeda1999,Marigo2007}, resulting in a higher SN Ia rate \citep{Kistler2013}, although note that \citet{Kistler2013} did not find any evidence for this in the data. \citet{Graur2017a} also found that evolution of SN Ia rates with metallicity is consistent with a DTD of the form  $\tau^{-1}$.

@@

Alternatively, low- and high-stretch SNe could form via identical evolutionary channels but with a different onset time since the episode of star-formation. In this scenario, he predominantly high-stretch "prompt" SNe begin exploding as soon as the WDs are formed (nominally 40 Myr) whereas the mainly low-stretch "tardy" SNe only begin exploding after an extended period of time of the order 1 Gyr. The time dependence of both populations would follow the fiducial $\tau^{-1}$ distribution.
Such a scenario may be explained by the sub-$M_{\mathrm{Ch}}$ double-detonation paradigm \citep[e.g.][]{Sim2010,Blondin2017,Shen2017}. Sub-$M_{\mathrm{Ch}}$ SNe Ia typically involve the merger of two WDs and the SN luminosity can be correlated with the mass of the primary WD; a correlation between primary WD mass and age leads to different DTDs of low and high-stretch SNe and hence the different rates observed in this work. Such an evolution of light curve properties with age has recently been seen across the full range of "normal" SNe Ia in 1-dimensional simulations by \citet{Shen2021} and is a promising avenue for further investigation although we note that they struggle to reproduce observed light curves at later times ($t\gtrsim30$~d; \citealt{Shen2021,Gronow2021}).

@@

The SN Ia rates generated by the posterior predictive checks are shown in Fig. \ref{fig:rate_fitted_split_x1_2c_DTD}. The best fitting values are somewhat consistent with one another, with $t_1 =0.69^{+0.07}_{-0.13}$~Gyr and $t_2 = 0.81^{+0.13}_{-0.10}$~Gyr, corresponding to a relatively sharp transition between the two populations. Our $t_1$ and $t_2$ are reminiscent of the transition found by \citet{Brandt2010} who found high-stretch SNe to be confined to delay times $\lesssim 0.4$~Gyr while low-stretch SNe occurred with delay times $\gtrsim 2.4$~Gyr. On the other hand, the transition found here appears to occur at much shorter delay times than the several Gyrs found by \citet{Rose2019}. These results are somewhat surprising, as we might expect that the transition is not so sharp but a gradual correlation of delay time with the average stretch of SNe caused by the age vs WD-mass distribution, as predicted by \citet{Shen2021}. However we note that the intrinsic stretch distribution may not be a simple Gaussian, but could instead be bimodal \citep{Scolnic2016,Popovic2021}. The correlation of progenitor age and WD mass does not easily reproduce such a bimodal distribution, which may suggest that the sub-$M_{\mathrm{Ch}}$ channel cannot account for all SNe Ia. We also note that our method models the average SFH of galaxies and that the DTD is a statistical distribution: our results do not rule out high-stretch SNe arising from old progenitors, but suggest that such a scenario is unlikely.
